% =============================================================================
% main.tex — Zero-Shot Text Classification with Locally-Hosted Large Language Models
% Authors: Luigi Palumbo, Mengting Yu, Carolina Camassa
% =============================================================================
\documentclass[11pt,a4paper]{article}

% ---------- Packages ----------
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage[margin=2.5cm]{geometry}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{enumitem}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{float}
\usepackage{amsmath,amssymb}

% ---------- Bibliography ----------
\usepackage[style=apa,backend=biber,natbib=true]{biblatex}
\addbibresource{references.bib}

% ---------- Hyperref setup ----------
\hypersetup{
  colorlinks=true,
  linkcolor=blue!70!black,
  citecolor=blue!70!black,
  urlcolor=blue!60!black,
}

% ---------- Listings setup ----------
\lstset{
  language=Python,
  basicstyle=\ttfamily\small,
  keywordstyle=\color{blue!70!black}\bfseries,
  commentstyle=\color{green!50!black}\itshape,
  stringstyle=\color{red!70!black},
  breaklines=true,
  frame=single,
  numbers=left,
  numberstyle=\tiny\color{gray},
  backgroundcolor=\color{gray!5},
}

% ---------- Title ----------
\title{
  \textbf{Zero-Shot Text Classification with Locally-Hosted\\
  Large Language Models: The \texttt{ollama-classifier} Library}
}
\author{
  Luigi Palumbo\thanks{Corresponding author.}
  \and Mengting Yu
  \and Carolina Camassa
}
\date{\today}

% =============================================================================
\begin{document}

\maketitle

% ---------- Abstract ----------
\begin{abstract}
  Large Language Models (LLMs) have emerged as powerful zero-shot classifiers,
  yet their adoption in domains that demand data privacy, cost control, and
  offline operation remains limited by reliance on proprietary cloud APIs.
  We present \texttt{ollama-classifier}, an open-source Python library that
  enables zero-shot and few-shot text classification using locally-hosted LLMs
  through the Ollama runtime. The library provides constrained output generation,
  confidence scoring, batch processing, and integration with multiple inference
  backends (Ollama, vLLM, SGLang, llama.cpp), making LLM-based classification
  accessible without fine-tuning, API keys, or internet connectivity.
  A companion desktop GUI built with Flet allows non-technical users to leverage
  retrieval-augmented classification through an intuitive workflow.
  We benchmark the library against BART-large-MNLI on a COICOP product
  classification task, demonstrating that a compact 3B-parameter model served
  locally substantially outperforms the NLI-based baseline.
\end{abstract}

\noindent\textbf{Keywords:} zero-shot classification, large language models,
local inference, retrieval-augmented classification, Ollama, open-source software

% =============================================================================
\section{Introduction}
\label{sec:introduction}
% =============================================================================

The advent of Large Language Models (LLMs) has fundamentally transformed
natural language processing, enabling remarkable zero-shot and few-shot
performance across a wide range of tasks \citep{brown2020language,
radford2019gpt2, zhao2023survey}. Among these tasks, text classification---the
assignment of predefined categorical labels to free-form text---stands out as a
particularly impactful application, with use cases spanning sentiment analysis
\citep{wang2024sentiment}, economic taxonomy coding, and product categorization
\citep{nunes2025rac}.

Traditionally, text classification has relied on supervised machine learning
pipelines that require substantial labeled training data, feature engineering,
and model fine-tuning. While transformer-based architectures such as BERT
\citep{berki2025nlp} have improved classification accuracy, they remain
data-intensive and offer limited adaptability to new or evolving taxonomies
without costly retraining. The emergence of LLMs has challenged this paradigm:
models such as GPT-3 \citep{brown2020language} and GPT-4 \citep{openai2023gpt4}
have demonstrated that generative models can perform classification through
in-context learning, eliminating the need for task-specific training data.

However, leveraging these capabilities in production environments---particularly
in sensitive domains such as official statistics, healthcare, and
finance---introduces significant challenges. Proprietary cloud-based APIs incur
recurring costs, require internet connectivity, and raise data privacy concerns
when processing confidential or regulated information. These constraints limit
the applicability of LLM-based classification in settings where data sovereignty
and operational independence are paramount.

The recent proliferation of open-source LLMs---including LLaMA
\citep{touvron2023llama, meta2024llama3}, Mistral
\citep{jiang2023mistral}, Qwen \citep{yang2024qwen2}, and DeepSeek-R1
\citep{deepseek2025r1}---combined with lightweight serving frameworks such as
Ollama \citep{ollama}, has made it feasible to run capable language models on
consumer hardware. This shift creates an opportunity to build classification
tools that combine the zero-shot reasoning power of LLMs with the privacy,
cost-effectiveness, and offline capability of local inference.

We present \texttt{ollama-classifier}, an open-source Python library that
realizes this opportunity by providing a unified interface for zero-shot and
few-shot text classification using locally-hosted LLMs. The library is designed
around three core principles:

\begin{enumerate}[leftmargin=*]
  \item \textbf{No fine-tuning required.} Classification is performed through
        prompting and retrieval augmentation, leveraging the model's pre-trained
        knowledge without any task-specific training.
  \item \textbf{Local and private.} All inference runs on the user's hardware
        through the Ollama runtime or compatible backends, ensuring that no
        data leaves the local machine.
  \item \textbf{Accessible to non-technical users.} A companion GUI application
        provides a point-and-click interface for the full classification
        workflow, abstracting away the complexity of prompt engineering,
        backend configuration, and batch processing.
\end{enumerate}

The remainder of this paper is organized as follows.
Section~\ref{sec:related} reviews related work on LLM-based classification and
retrieval augmentation.
Section~\ref{sec:background} provides background on the enabling technologies.
Section~\ref{sec:library} describes the \texttt{ollama-classifier} library's
architecture and API.
Section~\ref{sec:gui} introduces the companion GUI application.
Section~\ref{sec:application} illustrates a real-world application to product
classification and presents a benchmark evaluation comparing
\texttt{ollama-classifier} against a BART-large-MNLI zero-shot baseline.
Section~\ref{sec:discussion} discusses design decisions and limitations, and
Section~\ref{sec:conclusion} concludes with directions for future work.

% =============================================================================
\section{Related Work}
\label{sec:related}
% =============================================================================

\subsection{LLMs as Classifiers}

The use of generative language models as classifiers has its roots in the
observation that autoregressive LLMs can perform discriminative tasks through
careful prompt design. \citet{radford2019gpt2} first demonstrated that GPT-2
could perform zero-shot task transfer without explicit supervision, establishing
the paradigm of leveraging generative pre-training for discriminative
downstream tasks. This capability was dramatically scaled by GPT-3
\citep{brown2020language}, which showed that few-shot in-context learning could
achieve competitive performance on classification benchmarks without any
gradient updates.

Subsequent work has explored the boundaries of this approach.
\citet{liu2023preliminary} provide a systematic survey of prompting methods,
cataloguing strategies for instructing LLMs to perform classification through
template-based prompts. \citet{mu2024prompt} investigate how prompt complexity
affects zero-shot classification performance, finding that well-structured
prompts with explicit instructions and output constraints significantly improve
accuracy. \citet{gilardi2023chatgpt} demonstrate that ChatGPT can match or
exceed crowd-worker quality on text annotation tasks, providing empirical
support for using LLMs as practical classification tools.

\subsection{Retrieval-Augmented Classification}

Retrieval-Augmented Generation (RAG), introduced by \citet{lewis2020retrieval},
grounds LLM outputs in externally retrieved evidence, improving both factuality
and stability. This paradigm has been adapted to classification tasks through
Retrieval-Augmented Classification (RAC). \citet{yu2023retrieval} demonstrate
that combining nearest-neighbor retrieval with few-shot prompting improves text
classification without fine-tuning. \citet{abdullahi2024retrieval} extend this
to the zero-shot setting, showing that retrieved exemplars can enable
classification into arbitrary taxonomies without training data.

Most recently, \citet{nunes2025rac} apply RAC to COICOP 2018 product
classification, constructing a synthetic knowledge base of 34,451 exemplars
and achieving up to 0.80 accuracy at the finest taxonomy level. The
\texttt{ollama-classifier} library builds directly on this line of research,
operationalizing the RAC framework as a general-purpose, reusable Python tool.

\subsection{Existing Classification Libraries}

Several open-source projects address LLM-based text classification.
\texttt{llmclassifier} \citep{cvxtz2023llmclassifier} provides a Python
framework for building classification pipelines with LLMs, but targets API-based
models and does not support local inference. \texttt{llmClassificR}
\citep{golino2023llmclassificr} offers similar functionality in R, again
relying on cloud-hosted models.

\texttt{ollama-classifier} differentiates itself by (i) focusing exclusively on
local inference through Ollama and compatible backends, (ii) providing
constrained output generation that forces the model to return only valid class
labels, (iii) implementing confidence scoring through multi-call softmax
evaluation, and (iv) offering a dedicated GUI application for non-technical
users---features not found together in any existing library.

% =============================================================================
\section{Background}
\label{sec:background}
% =============================================================================

\subsection{The Ollama Runtime}

Ollama \citep{ollama} is an open-source framework for running LLMs locally on
consumer hardware. It provides a lightweight model server with a REST API,
model management (pull, list, delete), and support for quantized inference
that enables running models ranging from 1B to 70B+ parameters on standard
workstations.

The Ollama Python library \citep{ollama_python} provides a convenient client
interface, handling API communication and structured outputs. The
\texttt{ollama-classifier} library wraps this client to offer a higher-level
classification API, abstracting away prompt construction, output parsing, and
confidence computation.

\subsection{Inference Backends}

While Ollama serves as the default backend, \texttt{ollama-classifier}
supports multiple inference engines through a unified interface: Ollama for
consumer hardware and single-user workflows; vLLM for high-throughput batch
inference; SGLang for low-latency serving; and llama.cpp for resource-constrained
CPU and mixed CPU/GPU environments. All backends share a common configuration
interface, allowing users to switch between them without modifying application
code.

\subsection{Embedding Models for Retrieval}

The retrieval component of RAC relies on embedding models that convert text
into dense vector representations. SentenceTransformers
\citep{reimers2019sentence} provides the foundational Siamese BERT
architecture for producing semantically meaningful sentence embeddings. The E5
family \citep{wang2022text} offers multilingual semantic similarity across
100 languages, while the GTE family \citep{chen2024gte} provides
instruction-tuned embeddings with strong multilingual performance. In the RAC
evaluation by \citet{nunes2025rac}, the GTE models achieved the highest
retrieval presence (90.7\%), confirming their effectiveness for
classification-oriented retrieval tasks.

% =============================================================================
\section{The \texttt{ollama-classifier} Library}
\label{sec:library}
% =============================================================================

\subsection{Design Philosophy}

\texttt{ollama-classifier} is designed as a lightweight, modular Python library
that enables users to perform text classification with LLMs through a minimal
API surface. The design follows these principles:

\begin{enumerate}[leftmargin=*]
  \item \textbf{Simplicity}: A classification call requires only a text input
        and a list of candidate labels. All complexity---prompt construction,
        output parsing, backend communication---is handled internally.
  \item \textbf{Flexibility}: Users can customize the system prompt, provide
        label descriptions, choose between multiple classification strategies,
        and switch inference backends transparently.
  \item \textbf{Reliability}: Constrained output generation ensures that the
        model returns only valid class labels, preventing hallucinated or
        malformed predictions. Confidence scoring enables triage between
        high-confidence auto-accepted labels and low-confidence cases
        requiring human review.
  \item \textbf{Scalability}: Batch and async APIs enable processing large
        datasets efficiently, with concurrent requests supported across all
        backends.
\end{enumerate}

\subsection{Core API}

The library exposes two primary classifier classes:
\texttt{OllamaClassifier} (Ollama-specific) and \texttt{LLMClassifier}
(backend-agnostic). Both share an identical API surface, shown in
Table~\ref{tab:api}.

\begin{table}[htbp]
  \centering
  \caption{Core methods of the \texttt{ollama-classifier} API. Each method
           has an asynchronous counterpart prefixed with \texttt{a}.}
  \label{tab:api}
  \begin{tabular}{lp{8cm}}
    \toprule
    \textbf{Method} & \textbf{Description} \\
    \midrule
    \texttt{generate()} & Constrained output only; returns the predicted label
                          without confidence scores (fastest method). \\
    \texttt{score()}    & Multi-call evaluation with softmax; makes $N$ API
                          calls for $N$ choices to compute calibrated
                          probabilities. \\
    \texttt{classify()} & Full classification with confidence scores; combines
                          generation with probability estimation. \\
    \texttt{batch\_generate()} & Batch constrained output for multiple texts. \\
    \texttt{batch\_score()}    & Batch scoring for multiple texts. \\
    \texttt{batch\_classify()} & Batch classification with confidence scores. \\
    \bottomrule
  \end{tabular}
\end{table}

\subsection{Classification Strategies}

\texttt{ollama-classifier} supports three classification strategies, each
offering a different trade-off between speed and confidence calibration:

\begin{enumerate}[leftmargin=*]
  \item \textbf{Generate only} (\texttt{generate}): The model receives a single
        prompt containing the text and candidate labels, and returns one
        predicted label. This is the fastest strategy, suitable when speed is
        critical and confidence estimation is not required.

  \item \textbf{Multi-call scoring} (\texttt{score}): The model evaluates each
        candidate label independently through $N$ separate API calls, producing
        raw log-probability scores. These scores are then normalized using a
        softmax function to produce a calibrated probability distribution over
        all classes:
        \[
          \text{P}(y_i \mid x) = \frac{\exp(s_i)}{\sum_{j=1}^{N} \exp(s_j)},
        \]
        where $s_i$ is the log-probability score for class $y_i$ and $N$ is the
        number of candidate classes.

  \item \textbf{Full classification} (\texttt{classify}): Combines constrained
        generation with confidence scoring, returning both the predicted label
        and the probability distribution over all candidates. This strategy
        is recommended for applications that require both a prediction and a
        measure of certainty for triage purposes.
\end{enumerate}

\subsection{Constrained Output Generation}

A key design feature of \texttt{ollama-classifier} is the enforcement of
constrained output. The system prompt instructs the model to return only the
exact text of one of the provided candidate labels, preventing hallucinated
categories, partial responses, or verbose explanations. When supported by the
backend, JSON schema constraints are applied to structurally guarantee valid
output. For backends that do not support schema constraints (e.g., standard
Ollama), the prompt-based approach ensures reliable output formatting through
careful instruction design informed by the prompting literature
\citep{liu2023preliminary, mu2024prompt}.

\subsection{Confidence Scoring and Human-in-the-Loop}

The confidence scoring mechanism enables a human-in-the-loop workflow
\citep{schroeder2025just, hakimi2026humans}. High-confidence predictions can
be automatically accepted, while low-confidence predictions are flagged for
human review. \citet{vasquez2026trust} emphasize the importance of calibrating
LLM confidence scores for automated assessment, and the multi-call softmax
approach implemented in \texttt{ollama-classifier} provides a principled
foundation for such calibration.

The library also supports an ``opt-out'' mechanism, where an additional
``none of the above'' option is included among the candidate labels. This
allows the model to abstain from classification when none of the provided
options are appropriate, reducing the risk of forcing incorrect labels
\citep{nunes2025rac}. In the evaluation by \citet{nunes2025rac}, the
DeepSeek-R1 models demonstrated particularly well-calibrated opt-out behavior,
with correct abstention rates exceeding 85\%.

\subsection{Usage Example}

Listing~\ref{lst:usage} demonstrates a basic classification workflow using
\texttt{ollama-classifier}. The example shows how to initialize the classifier,
define candidate labels with descriptions, and classify a product description.

\begin{lstlisting}[caption={Basic usage of \texttt{ollama-classifier} for
  zero-shot product classification.},
  label={lst:usage}]
from ollama_classifier import OllamaClassifier

classifier = OllamaClassifier(model="llama3.2")

choices = {
    "Food and non-alcoholic beverages":
        "Edible items, fresh or packaged food, drinks",
    "Clothing and footwear":
        "Garments, shoes, accessories for personal use",
    "Electronics":
        "Consumer electronic devices and accessories",
}

result = classifier.classify(
    text="Organic whole wheat pasta 500g",
    choices=choices,
)
print(result)  # ClassificationResult(label=..., scores={...})
\end{lstlisting}

% =============================================================================
\section{Graphical User Interface}
\label{sec:gui}
% =============================================================================

While the Python library provides a powerful programmatic interface, many
potential users of LLM-based classification---domain experts, statisticians,
data officers, and analysts---may lack programming expertise.
\citet{amershi2019software} identify the gap between ML models and end-users
as a critical barrier to adoption, \citet{he2021automl} demonstrate that
automating ML workflows significantly lowers this barrier, and
\citet{lazaridou2023ai} argue for democratizing access to AI tools for domain
specialists.

To address this challenge, we developed \texttt{ollama-classifier-gui}, a
companion desktop application that provides a point-and-click interface for
the full classification workflow. The application is built with Flet, a
Python UI framework that enables cross-platform desktop applications from a
single codebase.

\subsection{Architecture}

The GUI application wraps the \texttt{ollama-classifier} library as a backend
engine, exposing its full functionality through an intuitive visual interface.
The application supports the same multiple inference backends as the library
(Ollama, vLLM, SGLang, llama.cpp), with all backend configuration managed
through a dedicated settings panel. API keys, where required, are stored using
the operating system's native secure storage mechanism (Keychain on macOS,
Credential Manager on Windows, libsecret on Linux), ensuring that sensitive
credentials are never stored in plain text.

\subsection{Four-Tab Workflow}

The application is structured around a four-tab workflow that mirrors the
classification pipeline: a \textbf{Settings} tab for configuring the inference
backend, model, and connection parameters; a \textbf{Data Input} tab for
loading text data from CSV or JSON files with column mapping and preview; a
\textbf{Schema} tab for defining candidate labels and descriptions, selecting
the classification strategy, and configuring custom system prompts; and a
\textbf{Results} tab for viewing predicted labels and confidence scores in a
structured table, with export to CSV or JSON. This workflow enables
non-technical users to perform classification without encountering code or
command-line interfaces.

\subsection{Significance}

The GUI application serves a critical role in the \texttt{ollama-classifier}
ecosystem by bridging the gap between the technical capabilities of modern
LLMs and the operational needs of domain practitioners. National Statistical
Institutes, research teams, and government agencies that need to classify
large volumes of text data---such as web-scraped product descriptions for
Consumer Price Index compilation \citep{cavallo2013online, cavallo2017online,
nunes2025rac}---can now leverage state-of-the-art LLM classification without
requiring software engineering expertise.

% =============================================================================
\section{Application to Product Classification}
\label{sec:application}
% =============================================================================

To illustrate the practical utility of \texttt{ollama-classifier}, we describe
its application to the classification of consumer products into the COICOP
2018 taxonomy \citep{un2018coicop}, a task of direct relevance to the
compilation of Consumer Price Indices by National Statistical Institutes.

\subsection{The Classification Challenge}

COICOP 2018 is an international standard developed by the United Nations to
categorize household expenditures by function. The taxonomy comprises a
four-level hierarchical structure, from broad divisions (e.g., ``Food and
non-alcoholic beverages'') down to fine-grained subclasses (e.g., ``Other
 edible products''), encompassing 294 subclasses at the most granular level.
Classifying web-scraped product descriptions into this taxonomy presents
several challenges: high linguistic variability across retailers and countries,
strong class imbalance, textual overlap between semantically similar
subclasses, and the need for multilingual coverage.

\subsection{Retrieval-Augmented Classification Pipeline}

\texttt{ollama-classifier} supports a full retrieval-augmented classification
pipeline. The workflow proceeds in three stages:

\begin{enumerate}[leftmargin=*]
  \item \textbf{Knowledge base construction}: Synthetic exemplars are generated
        for each COICOP subclass using LLM prompts informed by official
        definitions, then embedded using models such as GTE \citep{chen2024gte}
        or E5 \citep{wang2022text}.

  \item \textbf{Retrieval}: Given a product description, the system retrieves
        the top-$k$ most semantically similar exemplars and their associated
        COICOP labels.

  \item \textbf{Generative classification}: The retrieved context is
        incorporated into the classification prompt, and the LLM performs the
        final classification by reasoning over both the input text and the
        retrieved evidence.
\end{enumerate}

This pipeline, evaluated by \citet{nunes2025rac}, achieves up to 0.80 accuracy
at the finest COICOP level using the classification-by-definition approach with
Qwen2.5-32B \citep{yang2024qwen2}. Notably, even compact models such as
Mistral 7B \citep{jiang2023mistral} achieve strong overall performance (63.8\%
accuracy at Level 4), demonstrating that effective classification does not
require the largest available models.

\subsection{Deployment Scenarios}

\texttt{ollama-classifier} supports two primary deployment scenarios for this
application:

\begin{itemize}[leftmargin=*]
  \item \textbf{Programmatic pipeline}: Data scientists and engineers can
        integrate \texttt{ollama-classifier} into automated data processing
        pipelines, using the batch and async APIs to classify large volumes of
        products efficiently.

  \item \textbf{GUI-assisted workflow}: Non-technical users can load product
        datasets, define or import COICOP subclass labels, configure the
        classification parameters, and review results through the GUI
        application, with the option to export classified data for downstream
        analysis.
\end{itemize}

In both scenarios, the local inference model ensures that product data---which
may include commercially sensitive pricing information---never leaves the
user's machine.

\subsection{Experimental Evaluation}
\label{sec:evaluation}

To assess the practical performance of \texttt{ollama-classifier} on a
fine-grained product classification task, we conducted a benchmark comparison
against BART-large-MNLI \citep{lewis2019bart, williams2018mnli}, a widely-used
zero-shot classifier that frames text classification as natural language
inference (NLI).

\subsubsection{Setup}

We evaluated both classifiers on a manually-labeled dataset of 637 consumer
products drawn from COICOP Division 01.2 (Non-alcoholic beverages), sourced
from Zenodo. The dataset spans seven subclasses at the most granular level of
the COICOP 2018 hierarchy, with class sizes ranging from 15 to 189 products.
The candidate labels for each classifier are the seven subclass names (e.g.,
``Coffee and coffee substitutes,'' ``Water,'' ``Soft drinks''), optionally
accompanied by their official COICOP definitions and an opt-out option
(``None of the above'').

The baseline classifier is BART-large-MNLI \citep{lewis2019bart}, a 400M-parameter
seq2seq model fine-tuned on the MultiNLI corpus \citep{williams2018mnli}. BART
performs zero-shot classification by computing the entailment probability
between the input text (premise) and a hypothesis of the form ``This text is
about \textit{label}.'' Inference runs on CPU using the HuggingFace
\texttt{transformers} library.

The treatment classifier uses \texttt{ollama-classifier} with Qwen2.5 3B-Instruct
\citep{yang2024qwen2} served through Ollama \citep{ollama} on a local machine.
Classification uses the \texttt{classify} method, which combines
constrained output generation with multi-call confidence scoring to produce
both a predicted label and calibrated probability estimates over all
candidate classes.

We evaluate six variations: two for BART (names only, names with opt-out) and
four for \texttt{ollama-classifier} (names only, names with opt-out, names with
definitions, names with definitions and opt-out). This design allows us to
isolate the effects of (i) the underlying model, (ii) the opt-out mechanism,
and (iii) the inclusion of subclass definitions as additional context.

We report overall accuracy, macro-averaged precision, recall, and F1-score,
the percentage of items classified (excluding opt-outs), accuracy on classified
items only, and wall-clock processing time.

\subsubsection{Results}

Table~\ref{tab:benchmark} summarizes the results across all six variations.

\begin{table}[htbp]
  \centering
  \caption{Zero-shot classification results on COICOP Division 01.2
           (637 products, 7 subclasses). BART uses NLI-based zero-shot
           classification \citep{lewis2019bart}; Ollama uses the
           \texttt{classify} method (constrained generation with multi-call
           confidence scoring) with Qwen2.5 3B-Instruct \citep{yang2024qwen2}.
           Macro-averaged metrics are computed over the seven subclasses.
           Wall-clock time is reported in seconds.}
  \label{tab:benchmark}
  \begin{tabular}{lcccccc}
    \toprule
    \textbf{Variation} & \textbf{Acc.} & \textbf{Prec.} & \textbf{Rec.}
    & \textbf{F1} & \textbf{Classif.} & \textbf{Time} \\
    & (\%) & (macro) & (macro) & (macro) & (\%) & (s) \\
    \midrule
    BART (names only)         & 40.8 & 0.548 & 0.485 & 0.445 & ---   & 421.3 \\
    BART (names + opt-out)    & 39.4 & 0.549 & 0.495 & 0.454 & 94.2  & 466.3 \\
    \addlinespace
    Ollama (names only)       & 74.3 & 0.704 & 0.700 & 0.689 & ---   & 1814.4 \\
    Ollama (names + opt-out)  & 71.9 & 0.683 & 0.704 & 0.678 & 96.4  & 2043.3 \\
    Ollama (names + desc.)    & 70.8 & 0.684 & 0.704 & 0.677 & ---   & 1873.0 \\
    Ollama (desc. + opt-out)  & 69.1 & 0.692 & 0.719 & 0.685 & 93.7  & 2106.8 \\
    \bottomrule
  \end{tabular}
\end{table}

\textbf{Model comparison.}
\texttt{ollama-classifier} with Qwen2.5 3B-Instruct substantially outperforms
BART-large-MNLI across all configurations. The best Ollama variation (names
only, no opt-out) achieves 74.3\% overall accuracy compared to 40.8\% for the
best BART variation, representing a 33.5 percentage-point improvement. This
gap is consistent across opt-out configurations, indicating that the advantage
stems from the generative model's superior ability to reason over fine-grained
semantic distinctions rather than from the opt-out mechanism itself.

\textbf{Opt-out behavior.}
When the opt-out option is enabled, BART opts out of 37 products (5.8\%),
while Ollama opts out of 23 products (3.6\%). Among classified items,
Ollama's accuracy rises from 74.3\% to 74.6\% with opt-out enabled, suggesting
that the model correctly identifies its least-confident predictions. BART's
accuracy on classified items similarly improves from 40.8\% to 41.8\%, but the
magnitude of improvement is far smaller, consistent with BART's lower overall
discriminative capability.

\textbf{Effect of subclass definitions.}
Including official COICOP definitions alongside subclass names does not improve
classification performance for Ollama. The names-only variation (74.3\%)
outperforms the names-with-descriptions variation (70.8\%) by 3.5 percentage
points, indicating that definitional context actively harms performance in this
domain. This suggests that
for Division 01.2---where subclass names such as ``Coffee and coffee
substitutes'' and ``Tea, m\'at\'e and other plant products for infusion'' are
already highly distinctive---additional definitional context provides marginal
benefit and may occasionally introduce noise. This finding aligns with the
results of \citet{nunes2025rac}, who report that classification-by-definition
is most beneficial at higher COICOP levels where class boundaries are less
clear.

\textbf{Confidence and calibration.}
Table~\ref{tab:confidence} summarizes the relationship between classifier
confidence and prediction correctness across all six variations. Both models
exhibit a statistically significant positive correlation between the confidence
score assigned to the predicted label and whether the prediction is correct
(all $p < 0.001$). BART's Pearson correlation coefficients range from $r =
0.347$ to $r = 0.351$, while Ollama's range from $r = 0.402$ to $r = 0.452$.

The two models differ markedly in their confidence distributions. Ollama's mean
confidence for correct predictions is 0.954--0.959 across variations, compared
to 0.509--0.532 for BART. Notably, Ollama's mean confidence for
\emph{incorrect} predictions (0.804--0.820) exceeds BART's mean confidence for
\emph{correct} predictions. This indicates that Ollama's probability
estimates are compressed into a narrow high-confidence band, making absolute
threshold-based triage less effective.

The mean score gap---the difference between the highest and second-highest class
probabilities---further illuminates this pattern. For Ollama, the mean gap on
incorrect predictions is 0.91--0.92, indicating that the model strongly prefers
the wrong label over the correct one. BART's gap on incorrect predictions is
0.44--0.46, reflecting a more diffuse probability distribution where the model
is less decisive. These findings suggest that confidence thresholds are
informative for both models, but finer-grained triage strategies---such as
examining the score gap or the probability assigned to the ground-truth
class---may be more effective for Ollama than a simple threshold on the
predicted label's confidence.

\begin{table}[htbp]
  \centering
  \caption{Confidence analysis across classification variations.
           ``Conf (correct)'' and ``Conf (incorrect)'' report the mean
           predicted-label confidence for correct and incorrect predictions,
           respectively. ``Corr.~$r$'' is the Pearson correlation between
           predicted-label confidence and correctness. ``Score gap'' is the
           mean difference between the highest and second-highest class
           probabilities.}
  \label{tab:confidence}
  \begin{tabular}{lcccc}
    \toprule
    \textbf{Variation} & \textbf{Conf.} & \textbf{Conf.} & \textbf{Corr.}
    & \textbf{Score} \\
    & \textbf{(correct)} & \textbf{(incorrect)} & $\boldsymbol{r}$
    & \textbf{gap} \\
    \midrule
    BART (names only)         & 0.532 & 0.414 & 0.347 & 0.462 \\
    BART (names + opt-out)    & 0.509 & 0.392 & 0.351 & 0.441 \\
    \addlinespace
    Ollama (names only)       & 0.954 & 0.820 & 0.402 & 0.919 \\
    Ollama (names + opt-out)  & 0.959 & 0.810 & 0.437 & 0.921 \\
    Ollama (names + desc.)    & 0.956 & 0.804 & 0.452 & 0.911 \\
    Ollama (desc. + opt-out)  & 0.954 & 0.815 & 0.422 & 0.918 \\
    \bottomrule
  \end{tabular}
\end{table}

\textbf{Practical implications.}
These results demonstrate that a compact 3B-parameter model served locally
through Ollama can match or exceed the zero-shot classification performance of
a dedicated NLI model on a fine-grained product taxonomy task---without
requiring fine-tuning, API access, or data to leave the user's machine. The
constrained output generation mechanism ensures that all predictions are valid
class labels, eliminating the post-processing required by unconstrained LLM
outputs. The opt-out mechanism provides a principled way to flag uncertain
predictions for human review, a critical feature for production deployment in
official statistics.

\subsubsection{Limitations of the Evaluation}

Several caveats apply. The evaluation covers only one COICOP division (01.2,
Non-alcoholic beverages), and performance may differ at other divisions where
class boundaries are more ambiguous. We evaluate a single Ollama model
(Qwen2.5 3B-Instruct); larger models may achieve higher accuracy, while smaller
models may fall below the observed threshold. The BART baseline uses CPU
inference, and GPU acceleration could narrow the runtime gap. Finally, we do
not include a fine-tuned classifier baseline; such a comparison would require
task-specific training data and is left to future work.

% =============================================================================
\section{Discussion}
\label{sec:discussion}
% =============================================================================

\subsection{Design Decisions}

Several design decisions distinguish \texttt{ollama-classifier} from
alternative approaches:

\textbf{Local-first inference.} By defaulting to Ollama for local model
serving, the library prioritizes privacy and cost-effectiveness. While
cloud-based APIs offer access to larger models, the trade-off is often
unacceptable in sensitive domains. The support for multiple backends (vLLM,
SGLang, llama.cpp) ensures that users can choose the inference engine that
best matches their hardware and deployment requirements.

\textbf{No fine-tuning.} The library deliberately avoids any fine-tuning
step, relying entirely on prompting and retrieval augmentation. This design
choice maximizes flexibility: users can switch models, add or remove classes,
and adapt to new domains without retraining. The trade-off is that
fine-tuned models may achieve higher accuracy on specific tasks, but at the
cost of reduced generality and increased computational requirements.

\textbf{Constrained output.} Enforcing valid class labels through prompt
design and, where available, JSON schema constraints is essential for
production deployment. Unconstrained LLM outputs are prone to hallucination,
verbosity, and formatting inconsistencies that would break downstream
processing pipelines.

\subsection{Limitations}

Several limitations should be acknowledged. First, the quality of
classification depends on the underlying model's capabilities. While recent
open-source models have improved dramatically, smaller models (below 3B
parameters) may struggle with complex or ambiguous classification tasks. The
results reported by \citet{nunes2025rac} show that models with fewer than
1.5B parameters achieve only marginal classification accuracy, suggesting a
practical lower bound for effective deployment.

Second, bias and fairness concerns apply to LLM-based classification
\citep{gallegos2024bias}. The models' pre-training data may encode societal
biases that affect classification outcomes, particularly when classifying
texts related to demographic, economic, or cultural categories. Users should
be aware of these risks and consider bias auditing tools
\citep{yuan2026biasahead} when deploying \texttt{ollama-classifier} in
sensitive applications.

Third, the multi-call scoring strategy trades computational cost for
calibrated confidence estimates. For classification tasks with many candidate
labels, the $N$-call approach may be prohibitively expensive, and users may
need to rely on the faster generate-only strategy.

Finally, while the GUI application significantly lowers the barrier to entry,
it currently operates as a single-user desktop application. Multi-user,
server-based deployments would require additional infrastructure that is
beyond the current scope.

% =============================================================================
\section{Conclusion}
\label{sec:conclusion}
% =============================================================================

We have presented \texttt{ollama-classifier}, an open-source Python library
that enables zero-shot and few-shot text classification using locally-hosted
Large Language Models. The library provides constrained output generation,
confidence scoring, batch processing, and multi-backend support, making
advanced LLM-based classification accessible without fine-tuning, API keys,
or internet connectivity.

The companion GUI application extends this accessibility to non-technical
users, providing an intuitive four-tab workflow for the complete classification
pipeline---from backend configuration and data loading to schema definition
and results review. This combination of a programmatic library and a graphical
interface positions \texttt{ollama-classifier} as a practical tool for
bridging the gap between the capabilities of modern LLMs and the operational
needs of domain practitioners in fields such as official statistics, economic
research, and data governance.

Future work will focus on expanding the retrieval-augmented classification
pipeline within the library, improving confidence calibration through
temperature scaling and conformal prediction methods, and developing a
web-based interface to support collaborative, multi-user classification
workflows.

% =============================================================================
% References
% =============================================================================
\printbibliography

\end{document}
